The state learner -- a super learner for right-censored data

### 2025-06-05 manuscript/manuscript-one.tex
tags suggestions and marginnotes

% Handling new lines
\setlength{\parskip}{1em}
\setlength{\parindent}{0em}

% Handling space after sections
\usepackage{titlesec}
\titlespacing*{\section}{0em}{2em}{0em}
\titlespacing*{\subsection}{0em}{2em}{0em}
\titlespacing*{\subsubsection}{0em}{2em}{0em}

@@

\begin{abstract}
  In survival analysis, prediction models are needed as
  stand-alone tools and in applications of causal inference to
  estimate nuisance parameters. The super learner is a machine
  learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. In
  right-censored data, the choice of the loss function and the
  estimation of the expected loss need careful consideration.
  We introduce the state learner, a new super learner for
  survival analysis, which simultaneously evaluates libraries
  of prediction models for the event of interest and the
  censoring distribution. The state learner places no
  restrictions on the algorithms that can be included in the
  library, works in the presence of competing risks, and does
  not require a single pre-specified estimator of the
  conditional censoring distribution. We establish an oracle
  inequality for the state learner and investigate its
  performance through numerical experiments. We illustrate the
  application of the state learner with prostate cancer data.
\end{abstract}

@@

Accurate risk prediction from time-to-event data is an
important problem in many areas of medical research and public
health. Super learning \citep{van2007super}, also known as
ensemble learning or stacked regression
\citep{wolpert1992stacked,breiman1996stacked}, provides a
powerful approach to this problem by combining multiple
candidate prediction models. In the context of survival
analysis, a super learner can for instance combine Cox
regression models and random survival forests \citep[][Section
8.4]{gerds2021medical}.

In this paper we propose the state learner, a new super
learner designed to handle the specific challenges of ensemble
learning with right-censored data. The state learner estimates
the performance of a library of candidate models for both the
event-time and censoring distributions simultaneously. This
joint estimation framework avoids the need for pre-specifying
a censoring model, which is a limitation of many existing
methods. Our approach also allows for full flexibility in the
choice of models in the library, contrasting with other
proposals that restrict the library to specific model classes
\citep{polley2011-sl-cens,golmakani2020super}. Furthermore,
unlike iterative approaches such as those in
\cite{han2021inverse} and \cite{westling2021inference}, we can
provide theoretical guarantees for the performance of the
state learner.

The state learner is based on the simple idea of considering
right-censored competing risks data as an artificial
multi-state system where censoring is included as a state of
its own. Candidate models are then assessed based on how well
they predict the state occupations of this system across time,
based on baseline covariate information. We demonstrate how to construct
a library for the state learner using existing survival
models, and how to obtain risk predictions from the resulting
ensemble. To analyse the theoretical properties of the state
learner we focus on the discrete super learner which picks the
model in the library with best estimated performance
\citep{van2007super}. We show that the discrete state learner
is consistent under natural conditions and prove a finite-sample
oracle inequality.

\subsection{Related work}
\label{sec:related-work}

Super learning requires a library of candidate models, or
\textit{learners}, and a loss function for evaluating
performance on hold-out samples. Machine learning based on
right-censored data commonly uses the partial log-likelihood
as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
This loss function is also suggested for super learning with
right-censored data by \cite{polley2011-sl-cens}, where data
is assumed to be observed in discrete time. However, the
partial log-likelihood loss does not work well as a measure of
performance in hold-out samples observed in continuous time.
The reason is that the partial log-likelihood loss assigns an
infinite value for any learner that predicts piece-wise
constant cumulative hazard functions when there are event
times in the test set which do not occur in the learning set.
This problem occurs with prominent survival learners including
the Kaplan-Meier estimator, random survival forests, and
semi-parametric Cox regression models, and these learners
cannot be included in the library of the super learner
proposed by \cite{polley2011-sl-cens}. When a proportional
hazards model is assumed, the baseline hazard function can be
profiled out of the likelihood \citep{cox1972regression}. The
cross-validated partial log-likelihood loss
\citep{verweij1993cross} has therefore been suggested as a
loss function for super learning by \cite{golmakani2020super}.
This choice of loss function restricts the library of learners
to include only Cox proportional hazards models, and hence
excludes many learners such as, e.g., random survival forests,
additive hazards models, and accelerated failure time models.

More recent work by \cite{han2021inverse} and
\cite{westling2021inference} circumvents the need to
pre-specify a censoring model by iterating between estimation
of the outcome and censoring models. However, this procedure
is in general not guaranteed to converge to the true
data-generating mechanism
\citep[][Appendix~A.4]{munch2024thesis}.

@@

We introduce our notation and framework in
Section~\ref{sec:framework}. Section~\ref{sec:super-learning}
defines general super learning for right-censored data. In
Section~\ref{sec:super-learner-simple} we define the state
learner, while Section~\ref{sec:theor-results-prop} provides
theoretical guarantees. In Section~\ref{sec:use-cases-state}
we demonstrate how the state learner can be used to obtain
risk predictions, and we briefly mention how it can be applied
in the context of causal inference.
Section~\ref{sec:numer-exper} reports the results of numerical
experiments, and Section~\ref{sec:real-data-appl} illustrates
the method on prostate cancer data. We conclude with a
discussion in Section~\ref{sec:discussion}. Proofs are
collected in Appendix~\ref{sec:proofs}. Code and an
implementation of the state learner are available at
\url{https://github.com/amnudn/statelearner}.

% A super learner is a machine learning algorithm that combines
% a finite set of learners into a meta learner by estimating
% prediction performance in hold-out samples using a
% pre-specified loss function \citep{van2007super}. When the aim
% is to make a prediction model, super learners combine strong
% learners, such as Cox regression models and random survival
% forests \citep[][Section 8.4]{gerds2021medical}. While the
% general idea of combining strong learners based on
% cross-validation stems from earlier work
% \citep{wolpert1992stacked,breiman1996stacked}, the name super
% learner is justified by an oracle inequality
% \citep{van2003unicv,vaart2006oracle}.

% We define the state learner, a new super learner for
% right-censored data, which simultaneously estimates the
% expected loss of learners of the event time distribution and
% the censoring distribution. Compared to the super learners of
% \cite{polley2011-sl-cens} and \cite{golmakani2020super}, the
% state learner proposed here poses no restrictions on the type
% of learners that can be included in the library. Compared to
% super learners that rely on inverse probability of censoring
% weighting, censoring unbiased transformations, or
% pseudo-values, the state learner does not require a
% pre-specified censoring distribution. Compared to the super
% learners by \cite{han2021inverse} and
% \cite{westling2021inference}, the state learner has
% theoretical guarantees which these methods do not have.

% The loss function underlying the state learner operates on
% coarsened data. In the right-censored survival setting the
% coarsened data consist of the minimum and the order of the
% censoring time and the event time as well as the baseline
% covariates. The state learner works by simultaneously
% assessing how well learners of the event time distribution and
% the censoring distribution predict the states of the coarsened
% data. To analyse the theoretical properties of the state
% learner we focus on the discrete super learner which combines
% the library of learners by picking the one that minimises the
% cross-validated loss \citep{van2007super}. In the presence of
% competing risks, our algorithm uses separate libraries of
% learners for the cumulative hazard functions for each of the
% competing risks and for the censoring distribution. We show
% that the oracle selector of the state learner is consistent if
% all libraries contain a consistent learner and prove a finite
% sample oracle inequality.

% The state learner algorithm can output a medical risk
% prediction model \citep{gerds2021medical} which predicts the
% probability of an event based on covariates in the presence of
% competing risks. Another important application, which
% motivated our contribution, is in targeted learning where
% conditional event probabilities occur as high-dimensional
% nuisance parameters which need to be estimated at a certain
% rate \citep{van2011targeted, rytgaard2021estimation,
%   rytgaard2022targeted}. The state learner is particularly
% well-suited for this application as it simultaneously provides
% estimates for both the outcome and the censoring, both of
% which are needed for targeted learning. For the asymptotic
% bias term of a targeted estimator, which uses the state
% learner to estimate nuisance parameters, we show that a second
% order product structure holds. We illustrate both applications
% of the state learner with prostate cancer data
% \citep{kattan2000pretreatment}.

@@

In a competing risk framework \citep{andersen2012statistical},
let \( T\) be a time to event variable, \(D\in\{1,2\}\) the
cause of the event, and $X \in \mathcal{X}$ a vector of
baseline covariates taking values in a bounded subset
\( \mathcal{X} \subset \R^p \), \( p\in\N \). Let
$\tau< \infty$ be the prediction horizon. We use
\( \mathcal{Q} \) to denote the collection of all probability
measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
such that \( (T, D, X) \sim Q \) for some unknown
\( Q \in \mathcal{Q} \). For \(j\in\{1,2\}\), the
cause-specific conditional cumulative hazard functions
\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow
\R_+ \) are defined as
\begin{equation*}
  % \label{eq:cum-haz}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*}
For ease of presentation we assume throughout that the map
\( t\mapsto \Lambda_j(t \mid x) \) is continuous for all
\( x \) and \( j \). All arguments carry over directly to the
general case. We denote by \(S\) the conditional event-free
survival function,
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterised by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)=& \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=1\}}}\\
                         &  \left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}_{\tau}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

We consider the right-censored setting in which we observe
\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$
for a right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$,
and \(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of
probability measures on the sample space
\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times
\mathcal{X}\) such that \(O \sim P \) for some unknown
\(P\in \mathcal{P}\). We assume that the event times and the
censoring times are conditionally independent given
covariates, \( T \independent C \mid X \). This implies that
any distribution \( P \in \mathcal{P} \) is characterised by a
distribution \( Q \in \mathcal{Q} \) and a conditional
cumulative hazard function for \( C \) given \( X \)
\citep[c.f.,][]{begun1983information,gill1997coarsening}. We
use \(\Gamma\in\mathcal{M}_{\tau}\) to denote the conditional
cumulative hazard function for censoring. For ease of
presentation we now also assume that
\( \Gamma(\blank \mid x) \) is continuous for all \( x \). We
let
\((t,x)\mapsto G(t \mid x)=\exp\left\{-\Gamma(t \mid
  x)\right\}\) denote the survival function of the conditional
censoring distribution. The distribution \( P \) is
characterised by
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) =& \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                            & \left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                            & \left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
    = & \left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                            & \left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
also have \(H\)-almost everywhere
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*}
We further assume that there exists \(\kappa<\infty\) such that
\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
Under these assumptions, the conditional cumulative hazard functions
\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M_{\tau}\).

\section{Super learning}
\label{sec:super-learning}

In survival analysis, a super learner can be used to estimate
a parameter $\theta \in \Theta$ which can be identified from
the observed data distribution \(P\in\mathcal P\). A super
learner typically estimates a function, and thus the parameter
space $\Theta$ is most often a class of functions. For
example, $\theta$ could be a conditional survival function or
a cumulative hazard function. In
Section~\ref{sec:super-learner-simple} we consider a specific
choice of parameter as target for our super learner, but in
this section we describe super learning for a general
parameter $\theta$ with values in some parameter space
$\Theta$.

As input to a super learner we need a data set
\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from
\( P \in \mathcal{P} \) and a collection of candidate models
or learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\)
is a map \( a \colon \mathcal{O}^n \rightarrow \Theta \) which
takes a data set as input and returns an estimate
$a(\data_n) \in \Theta$ of $\theta$.
% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
In what follows, we use the short-hand notation
\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
performance of \(a \in \mathcal{A}\) with a loss function
\(L\colon \Theta \times \mathcal{O} \rightarrow \R_+\) and
estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
cross-validation. Specifically, the expected loss of $a\in\mathcal A$
is estimated by splitting the data set $\data_n$ into $K$ disjoint
approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  % \label{eq:cv-risk-est}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
The oracle learner is defined as the learner that minimises the
expected loss under the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner
depend on the library of learners and on the number of folds \(K\),
and that the oracle learner is a function of the data and the unknown
data-generating distribution. However, these dependencies are
suppressed in the notation.

@@

The main idea of the state learner is to jointly use learners
for \( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and
the relations in equation~(\ref{eq:parametrizeP}), to learn a
feature of the observed data distribution \( P \). The
discrete state learner ranks a tuple of learners for the tuple
of the cumulative hazard functions
\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they
jointly model the observed data. To formally introduce the
state learner, we define the artificial multi-state process
\begin{equation*}
  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde
  D=2\} - \1\{\tilde{T} \leq t, \tilde D=0\},
  \quad \text{for} \quad t \in [0, \tau].
\end{equation*}
We use the phrase `artificial' because multi-state models
should not usually include censoring as a state.
At time \(t\), we observe that each individual is in one of four mutually
exclusive states: \( 0 \), \( 1 \), \( 2 \), or \( -1 \).
The conditional distribution of the process \( \eta(t) \) given
baseline covariates \( X \) is determined by the function
\begin{equation}
  \label{eq:F-def}
  F(t, k, x) = P(\eta(t) = k \mid X=x),
\end{equation}
for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and
\( x \in \mathcal{X} \). The function \( F \) describes the
conditional state occupation probabilities of the process
\(\eta\).

>>>>

% Handling new lines
\setlength{\parskip}{1em}
\setlength{\parindent}{0em}
\usepackage{marginnote}
% Handling space after sections
\usepackage{titlesec}
\titlespacing*{\section}{0em}{2em}{0em}
\titlespacing*{\subsection}{0em}{2em}{0em}
\titlespacing*{\subsubsection}{0em}{2em}{0em}

@@

\begin{abstract} Risk prediction models are widely used to guide
real-world decision-making in areas such as healthcare and economics,
and they also play a key role in estimating nuisance parameters in
semiparametric inference. The super learner is a machine learning
framework that combines a library of prediction algorithms into a
meta-learner using cross-validated loss. In the context of
right-censored data, careful consideration must be given to both the
choice of loss function and the estimation of expected loss. Moreover,
estimators such as inverse probability of censoring weighting (IPCW)
require accurate modeling and estimates of the censoring distribution.

We propose a novel approach to super learning for survival analysis
that jointly evaluates candidate learners for both the event time
distribution and the censoring distribution. Our method imposes no
restrictions on the algorithms included in the library, accommodates
competing risks, and does not rely on a single pre-specified estimator
of the censoring distribution. We establish an oracle inequality for
our method, assess its performance through simulation studies, and
demonstrate its practical utility using prostate cancer data.
\end{abstract}

@@

Accurately predicting risk from time-to-event data is a central
challenge in diverse research fields including epidemiology, economics
and weather forecasting with applications in clinical
decision making and policy interventions. For instance, in prostate
cancer management, clinicians often need to estimate a patient’s risk
of disease progression and mortality over time to make informed
decisions about treatment strategies such as active surveillance
versus immediate intervention. Reliable time-to-event risk predictions
can help tailor care to individual patients, avoid overtreatment, and
allocate healthcare resources more effectively. Super learning
\citep{van2007super}, also known as ensemble learning or stacked
regression \citep{wolpert1992stacked,breiman1996stacked}, provides a
powerful approach to this problem which by combining multiple
candidate prediction models reduces the risk of bias incurred by a
single potentially mispecified model. In survival analysis, a super
learner may combine a stack of Cox regression models with a stack of
random survival forests \citep[][Section 8.4]{gerds2021medical}. Such
a strategy has recently produced KDpredict
(\url{https://kdpredict.com/}) a model which jointly predicts the
risks of kidney failure and all-cause mortality at multiple time
horizons based on different sets of covariates
\citep{liu2024predicting}. To evaluate the prediction performance of
the learners the super learner behind KDpredict uses inverse
probability of censoring weighting (IPCW) where the censoring
distribution is estimated under the restrictive assumption that the
censoring distribution does not depend on the covariates. This is a
potential source of bias which is difficult to overcome with the
currently available methods.

In this paper we propose the state learner, a new super learner
designed to handle the specific challenges of ensemble learning with
right-censored data. The state learner simultaneously learns
prediction models for the event-time and censoring distributions. The
state learner is based on the simple idea of an artificial competing
risk model where censoring is included as a state of its
own. Candidate learners of the event time hazard and censoring hazard
functions are then assessed based on how well they predict the state
occupation probabilities of the artificial competing risk model across
time based on baseline covariate information. Our estimation framework
allows for competing risks, avoids restrictive assumptions on the
censoring distribution, and is also fully flexible with respect to the
choice of learners. The latter is in contrast to other proposals which
restrict the library of learners to specific model classes
\citep{polley2011-sl-cens,golmakani2020super}, see Section
\ref{sec:choice-of-loss-function}. Recent work by
\cite{han2021inverse} and \cite{westling2021inference} also
circumvents the need to pre-specify a censoring model by iterating
between estimation of the outcome and censoring models. However, this
procedure is in general not guaranteed to converge to the true
data-generating mechanism \citep[][Appendix~A.4]{munch2024thesis}. To
analyse the theoretical properties of the state learner we focus on
the discrete super learner which picks the model in the library with
best estimated performance \citep{van2007super}. Here we provide
theoretical guarantees for the performance of the state learner and in
particular show that the discrete state learner is consistent under
natural conditions and prove a finite-sample oracle inequality. We
demonstrate how to construct a library for the state learner using
common survival models, and how to obtain risk predictions from the
resulting ensemble.\marginnote{is the term cross-fitting correct?}
The state learner can be applied to produce a risk
prediction model and can also be used for cross-fitting in semiparametric inference.

@@

We introduce our notation and framework in
Section~\ref{sec:framework}. Section~\ref{sec:super-learning}
introduces loss-based super learning. In
Section~\ref{sec:super-learner-simple} we define the state learner,
while Section~\ref{sec:theor-results-prop} provides theoretical
guarantees. In Section~\ref{sec:use-cases-state} we demonstrate how
the state learner can be used to obtain risk predictions, and we
briefly mention how it can be applied in the context of causal
inference.  Section~\ref{sec:numer-exper} reports the results of
numerical experiments, and Section~\ref{sec:real-data-appl}
illustrates the method on prostate cancer data. We conclude with a
discussion in Section~\ref{sec:discussion}. Proofs are collected in
Appendix~\ref{sec:proofs}. Code and an implementation of the state
learner are available at \url{https://github.com/amnudn/statelearner}.

@@

In a competing risk framework \citep{andersen2012statistical}, let \(
T\) be a time to event variable, \(D\in\{1,2\}\) the cause of the
event, and $X \in \mathcal{X}$ a vector of baseline covariates taking
values in a bounded subset \( \mathcal{X} \subset \R^p \), \( p\in\N
\). Let $\tau< \infty$ be a fixed prediction horizon. We use \(
\mathcal{Q} \) to denote the collection of all probability measures on
\( [0,\tau] \times \{1,2\}\times \mathcal{X} \) such that \( (T, D, X)
\sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard
functions \( \Lambda_{j} \colon [0, \tau] \times \mathcal{X}
\rightarrow \R_+ \) are defined as
\begin{equation*}
  % \label{eq:cum-haz}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*} For ease of presentation we assume throughout that the
map \( t\mapsto \Lambda_j(t \mid x) \) is continuous for all \( x \)
and \( j \), however, 
all technical arguments extend naturally to the general case \citep{andersen2012statistical}.
The event-free survival function conditional on covariates is given by
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterised by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)=& \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=1\}}}\\
                         &  \left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}_{\tau}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

We consider the right-censored setting in which we observe \(O =
(\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and \(\tilde
D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability measures
on the sample space \(\mathcal{O} = [0, \tau] \times \{0, 1, 2\}
\times \mathcal{X}\) such that \(O \sim P \) for some unknown \(P\in
\mathcal{P}\). We assume that the event times and the censoring times
are conditionally independent given covariates, \( T \independent C
\mid X \). This implies that any distribution \( P \in \mathcal{P} \)
is characterised by a distribution \( Q \in \mathcal{Q} \) and a
conditional cumulative hazard function for \( C \) given \( X \)
\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
\(\Gamma\in\mathcal{M}_{\tau}\) to denote the cumulative hazard
function of the conditional censoring distribution given
covariates. For ease of presentation we assume that \(t\mapsto
\Gamma(t \mid x) \) is continuous for all \( x \). We let
\((t,x)\mapsto G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\)
denote the survival function of the conditional censoring
distribution. The distribution \( P \) is characterised by
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) =& \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                            & \left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                            & \left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
    = & \left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                            & \left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
also have \(H\)-almost everywhere
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*} We assume that there exists \(\kappa<\infty\) such
that \(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal
X\). Note that this implies that \(G(\tau- \mid x)\) is bounded away
from zero for almost all \(x\in\mathcal X\).  Under these assumptions,
the conditional cumulative hazard functions \(\Lambda_{j}\) and
\(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M_{\tau}\).

\section{Loss-based super learning}
\label{sec:super-learning}

\marginnote{formally super learning may work without loss function}
Loss-based super learning requires a library of candidate models, or
\textit{learners}, a cross-validation algorithm, and a loss function
for evaluating prediction performance on hold-out samples. In this
section we provide an abstract definition of loss-based super
learning.  As input to a super learner we need a data set \(
\data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in
\mathcal{P} \) and a collection of candidate models or learners
$\mathcal{A}$. Let \(\Theta\) be a suitable parameter space such as
\([0,1]\) for risk predictions, or the space of cadlag functions on
\([0,\tau]\) for predictions of survival functions.  Each learner \(a
\in \mathcal{A}\) is a map \( a \colon \mathcal{O}^n \rightarrow
\Theta \) which takes a data set as input and returns an estimate
$a(\data_n)$ of $\theta$.  \marginnote{we do not use Pf throughout but
also use integral and E notation!? would be good to reduce the amount
of abstract notation and settle for one way} In what follows, we use
the short-hand notation \(P[f] = \int f(o) P(\diff o) \).  Let
\(L\colon \Theta \times \mathcal{O}^n \rightarrow \R_+\) be a loss
function and denote the expected loss by \(P[L(a(\data_n), \blank)]\)
where the expectation is taken across all possible data sets of size
\(n\).

The expected loss of a learner is estimated by splitting the data set
$\data_n$ into $K$ disjoint approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  % \label{eq:cv-risk-est}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
The oracle learner is defined as the learner that minimises the
expected loss under the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner
depend on the library of learners and on the number of folds \(K\),
and that the oracle learner is a function of the data and the unknown
data-generating distribution.
% However, these dependencies are suppressed in the notation.

@@

The main idea of the state learner is to jointly use learners for \(
\Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
equation~(\ref{eq:parametrizeP}), to learn a feature of the observed
data distribution \( P \). The discrete state learner ranks a tuple of
learners for the tuple of the cumulative hazard functions \(
(\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model
the observed data. To formally introduce the state learner, we define
the process 
\begin{equation*}
  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde
  D=2\} - \1\{\tilde{T} \leq t, \tilde D=0\},
  \quad \text{for} \quad t \in [0, \tau]
\end{equation*} which takes values in \( \{-1,0,1,2\}\) which
represent four mutually exclusive states. Specifically value \( 0 \)
represents the state where the individual is still event-free and
uncensored, value \( 1\) the state where the event of interest has
occured, value \( 2\) the state where a competing risk has occured and
value \( -1\) the state where the observation is right censored.  The
state occupation probabilities given baseline covariates \( X \) are
given by the function\marginnote{cannot use k because k is used in
previous section}
\begin{equation}
  \label{eq:F-def}
  F(t, k, x) = P(\eta(t) = k \mid X=x),
\end{equation}
for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and
\( x \in \mathcal{X} \).
